---
title: "Assignment 3: Data Exploration"
author: "Key"
output: pdf_document
geometry: margin=2.54cm
---

# Total: 17 points

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 28 at 1:00 pm.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 

**2 points** 1/2 for wd, 1/2 for tidyverse, 1/2 for each csv
```{r, message = FALSE}
getwd()
library(tidyverse)

Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv")
Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv")
```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

**1 point** contains thoughtful answer

> Answer: e.g., identifies effects on both target and non-target species, dangers of colony collapse disorder in pollinators, food web effects

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

**1 point** contains thoughtful answer

> Answer: e.g., forest carbon balance, soil organic matter recharge, detritus portion of food web

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

**1 point** includes three pieces of relevant information from the user guide.

> Answer: 
*
*
*



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

**1/2 point**
```{r}
dim(Neonics)
```

6. Using the `summary` function, determine the most common effects that are studied. Why might these effects specifically be of interest?

**1 point** 1/2 for code, 1/2 for explanation
```{r}
summary(Neonics$Effect)
```

> Answer: most common: mortality and population. Since these are insecticides, we are interested in direct toxicological effects that affect survival at the individual and population levels.

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

**1 point** 1/2 for code, 1/2 for explanation

```{r}
summary(Neonics$Species.Common.Name)
```

> Answer: 5/6 are bees, and the remaining species is a wasp (same order as bees). These insects are of interest due to colony collapse disorder and pollinator decline. 

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

**1 point** 1/2 for code, 1/2 for explanation

```{r}
class(Neonics$Conc.1..Author.)
summary(Neonics$Conc.1..Author.)
```

> Answer: There are some letters and some characters other than numbers


## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 

**1 point**
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year))
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

**1 point** 1/2 for code, 1/2 for explanation
```{r}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location))
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Lab and field natural are most common with fairly equal counts prior to 2010 and then lab dominates after that.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

**1 point**
```{r}
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint))
```

> Answer: NOEL (No-observable-effect-level: highest dose (concentration) producing effects not significantly different from responses of controls)
> LOEL(Lowest-observable-effect-level: lowest dose (concentration) producing effects that were significantly different)

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 

**1.5 points** 1/2 for each
```{r}
class(Litter$collectDate)
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

**1 point** 1/2 for code, 1/2 for answer
```{r}
unique(Litter$plotID)
summary(Litter$plotID)
```

> Answer: 12 plots. Unique shows which levels are unique and how many, and summary shows the count for each level.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

**1 point**
```{r}
ggplot(Litter) +
  geom_bar(aes(x = functionalGroup))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 

**2 points** 1/2 for each plot, 1/2 for first answer, 1/2 for second answer
```{r}
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass))

ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: Violin plots are either very wide and short or very long and skinny, neither of which are effective ways to show the distribution

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles and mixed